Large Linear Multi-output Gaussian Process Learning for Time Series

### 2017-05-19 paper/paper.tex
targetted editing

Gaussian processes (GPs) are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \cite{williams1996gaussian}. GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes.

Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}. Efficient model estimation would enable researchers to quickly explore large spaces of parameterizations to find an appropriate one for their task.

% TODO: is this paragraph OK now?
% yes, except you need to cite more.
For $n$ input points, exact GP inference requires maintaining an $n^2$  matrix of covariances between response variables at each input and performing $O(n^3)$ inversions with that matrix \cite{williams1996gaussian}. Some single-output GP methods exploit structure in this matrix to reduce runtime \cite{msgp}. In the multi-output setting, the same structure does not exist. Approximations developed for multi-output methods instead reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \cite{nguyen2014collaborative}. The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure. We use this structure to avoid direct matrix inversion.

@@

A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}. With zero mean wlog and a prescribed covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$. Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \cite{williams1996gaussian}.
Predictive accuracy is sensitive to a particular parameterization of our kernel, and model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$. Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
\begin{align}
\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}};\;\; \bsa=K_{|\bsth}^{-1}\Ty.\label{llgradient}
\end{align}

@@

If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation. The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$th block $A_{ij}B$. We can represent the product by keeping representations of $A$ and $B$ separately, rather than the product. Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM. For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O(n^{3/P})$ \cite{gilboa2015scaling}.

@@

SKI abandons the inducing-point approach: instead of using an intrinsically sparse model, SKI approximates the original $K_{X,X}$ directly \cite{kiss-gp}. To do this efficiently, SKI relies on the differentiability of $K$. For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\cite{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$. The simultaneous interpolation $W\triangleq W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx WK_{U,U}W^\top$. $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$. SKI thus creates an $m$-rank approximation to $K_{X,X}$, reducing representation size even without relying on structure.

@@

With multiple outputs, non-differentiable cross-covariance interactions violate the assumptions of SKI's cubic interpolation. We show  a modification to SKI is compatible with the piecewise-differentiable LMC kernel. This approximation induces the previously-identified block-Toeplitz structure of the LMC kernel even for non-uniform inputs, which we use to accelerate GP estimation in the general case.

@@

For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$. Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$. Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.

@@

\citet{gibbs1996cient} describe the algorithm for GP model estimation in terms of only MVMs with the covariance matrix. In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG). Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
\begin{align}
  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}
\end{align}
For this approximation, the number of samples may be small, and estimates improve as the size of $K_{|\bsth}$ increases.
As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.

@@

For the rank-based \textsc{slfm} representation, let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average rank, $R\le D$, and re-write the kernel:
\begin{align*}
  K_{\TU,\TU}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
\end{align*}
Note $\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top$ is rank $1$. Under some re-indexing $q'\in[RQ]$, which flattens the double summation such that each $q'$ corresponds to a unique $(r, q)$, the term $\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q $ may be rewritten as
\begin{align*}
  \sum_{q'}\textbf{a}_{q'}\textbf{a}_{q'}^\top\otimes K_{q'} = \textbf{A}\blockdiag_{q'}\colv{K_{q'}}\textbf{A}^\top,
\end{align*}
where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}. Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz. Thus, \textsc{slfm} represents $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs take $O((QR + D)m\log m)$ time.

@@

We use the default COGP parameters\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}. The weather dataset was run on 1500 iterations, mini-batch size $1000$, $200$ interpolating points.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model. We tested LLGP models on different numbers of interpolating points $m = \{500, 1000\}$.

>>>>

Gaussian processes (GPs) are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \cite{williams1996gaussian}. GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes. Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}. Efficient model estimation would enable researchers to quickly explore large spaces of parameterizations to find an appropriate one for their task.

For $n$ input points, exact GP inference requires maintaining an $n^2$  matrix of covariances between response variables at each input and performing $O(n^3)$ inversions with that matrix \cite{williams1996gaussian}. Some single-output GP methods exploit structure in this matrix to reduce runtime \cite{msgp}. In the multi-output setting, the same structure does not exist. Approximations developed for multi-output methods instead reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \cite{nguyen2014collaborative}. The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure. Here, we exploit this structure to avoid direct matrix inversion.

@@

A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}. With zero mean without loss of generality and a prespecified covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$. Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \cite{williams1996gaussian}.
Predictive accuracy is sensitive to a particular parameterization of our kernel, and model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$. Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
\begin{align}
\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}};\;\; \bsa=K_{|\bsth}^{-1}\Ty.\label{llgradient}
\end{align}

@@

If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation \cite{gilboa2015scaling, cunningham2008fast}. The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$th block $A_{ij}B$. We can represent the product by keeping representations of $A$ and $B$ separately, rather than the product. Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM. For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O(n^{3/P})$ \cite{gilboa2015scaling}.

@@

SKI abandons the inducing-point approach: instead of using an intrinsically sparse model, SKI approximates the original $K_{X,X}$ directly \cite{kiss-gp}. To do this efficiently, SKI relies on the differentiability of $K$. For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\cite{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$. The simultaneous interpolation $W\triangleq W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx WK_{U,U}W^\top$. $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$. Even without relying on structure, SKI reduces the representation of $K_{X,X}$ to an $m$-rank matrix.

@@

With multiple outputs, non-differentiable cross-covariance interactions violate the assumptions of SKI's cubic interpolation. We show a modification to SKI is compatible with the piecewise-differentiable LMC kernel. This approximation induces the previously-identified block-Toeplitz structure of the LMC kernel, enabling acceleration of GP estimation for non-uniform inputs.

@@

For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$. Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$. Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as a gradient-only high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.

@@

\citet{gibbs1996cient} describe the algorithm for GP model estimation in terms of only MVMs with the covariance matrix. In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG). Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
\begin{align}
  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}.
\end{align}
This approximation improves as the size of $K_{|\bsth}$ increases, so, as in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.

@@

For the rank-based \textsc{slfm} representation, let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average rank, $R\le D$, and re-write the kernel:
\begin{align*}
  K_{\TU,\TU}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
\end{align*}
Note $\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top$ is rank $1$. Under some re-indexing $q'\in[RQ]$, which flattens the double summation such that each $q'$ corresponds to a unique $(r, q)$, the term $\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q $ may be rewritten as
\begin{align*}
  \sum_{q'}\textbf{a}_{q'}\textbf{a}_{q'}^\top\otimes K_{q'} = \textbf{A}\blockdiag_{q'}\colv{K_{q'}}\textbf{A}^\top,
\end{align*}
where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}. Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz. Thus, \textsc{slfm} represents $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.

@@

We use the default COGP parameters\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}. The weather dataset was run on 1500 iterations, mini-batch size $1000$, $200$ interpolating points.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model. We tested LLGP models on $m=\{500, 1000\}$ interpolating points.
